@$\lambda$: Conservatively Extending a Type System From Within

### 2014-02-28 att-icfp14/att-icfp14.tex
More icfp progress

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Typed programming languages are often described in fragments, each consisting of a small number of type constructors (often one) and associated (term-level) operator constructors. The simply typed lambda calculus (STLC), for example, consists of a single fragment containing a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors: $\lambda$, indexed by a type, and \verb|ap|, which we may think of as being indexed trivially. A fragment is often identified by the type constructor it is organized around, so the STLC might also be called $\mathcal{L}\{\rightarrow\}$, following the notational convention used by Harper \cite{pfpl} and others.

@@

If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e. parametric polymorphism), we might take pause. In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined via a technique similar to Church encodings \cite{reynolds}\todo{what to cite for this?}. While of substantial theoretical (and pedagogical) interest, this fact does not render irrelevant the construction of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$. Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, cautioned against using the existence of a weak encoding as an argument for the practicality of programming with the corresponding embedding directly \cite{Reynolds94anintroduction}: 
\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
\end{quote}

Finding a weak encoding of a language fragment in terms of a collection of others is useful if \emph{implementing the dynamic semantics correctly} is one's main interest, but adding new fragments directly to a language can make statically reasoning about programs more precise, support a more natural programming style and endow the language with a more favorable cost semantics.  
Consistent with this view, typed programming languages like ML expose, for example, both $n$-ary product types and record types, despite the fact that any particular product or record type is weakly definable using  polymorphic functions (or simply in terms of the other). The compiler for such a language, on the other hand, \emph{is} mainly concerned with implementing the dynamic semantics correctly, and indeed, compilers often use the fact that weak definability results are quite readily derivable to their advantage, translating the constructs of their external language (EL) to those of a much simpler internal language (IL) after typechecking is complete. %Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases needing consideration when implementing and verifying the compiler is a wise choice.

Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, we might now ask whether modern ``general-purpose'' languages like ML are fundamentally different. These languages have been used successfully for a wide variety of computing tasks, admitting many more strong encodings than a  calculus like $\mathcal{L}\{\rightarrow \forall\}$. It is certainly not our intention to argue against the notion that the particular fragments they have chosen to privilege occupy ``sweet spots'' in the design space. But in fact, situations continue to arise in both research and practice where the direct introduction of a new fragment is desirable because it can still only be weakly defined in terms of existing fragments, or a strong encoding is possible but impractically verbose or inefficient:
\begin{enumerate}
\item General-purpose abstractions continue to evolve and admit variants. There are many  variations on record types, for example: records with functional record update, prototypic delegation or specified field ordering (that is, labeled tuples) are either awkward to strongly define or cannot be. Similarly, sum types also admit variants (e.g. various forms of open sums). Even something as seemingly simple as a type-safe \verb|printf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.

\item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers. One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and a bounty of others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\item Foreign function interfaces (FFIs) that allow programmers to safely and naturally interact with libraries written in an existing language require enforcing the type system of the foreign language within the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
Safe FFIs generally require direct extensions to the language. For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{mlj}. For any other language, including  others on the JVM like Scala and the many languages that might be accessible via Java's native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
\end{enumerate}

These language fragments are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, a feature behind a compiler flag or, using tools like compiler generators, DSL frame\-works or  language workbenches. This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined. Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from language   composition operations, so even  fragmentary frameworks induce an unbounded, combinatoric collection of proof obligations that providers have little choice but to  burden clients with.% We will consider this further in our discussion of related work in Sec. \ref{related-work}.

These are not the sorts of problems usually faced by library providers. Well-designed languages preclude the  possibility of conflict between two libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers. For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without attending to burdensome ``link-time'' proof obligations.

We draw inspiration from this approach to design a mechanism that makes it possible to extend the static and dynamic semantics of an external portion of the language with new type and operator constructors from within. The semantics of the language as a whole constrains these extensions to maintain type safety and decidability of type checking and type equality. Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong encodings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively). The language  guarantees that they will be conserved in the  ``open world'' by enforcing  representation independence at extension boundaries.

Our mechanism essentially tasks the type-level language, in a controlled manner, with implementing portions of the logic that would normally be implemented in the front-end of the compiler. When typechecking a user-defined operator invocation, the semantics invoke a type-level function associated with the operator's constructor to \emph{compute} both a type and a translation to a fixed typed internal language (together called a \emph{derivate}). Language fragments that can be weakly defined in terms of the typed internal language can be made strongly definable in the external language by introducing, in this way, the necessary  constructors and static semantics. We call this an \emph{actively typed semantics}.

We will begin in Sec. \ref{atlam} by introducing a minimal calculus with an actively typed semantics, @$\lambda$. The external language builds in only the $\rightarrow$ fragment. We then use the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ and a fragment defining $n$-ary products as pedagogical examples to explain how extensions are written and motivate the constraints that the semantics must impose to make the guarantees that we claim. We then examine the semantics in greater detail and state and sketch proofs of strong safety and conservativity theorems in Sec. \ref{theory}. \todo{organization of rest of sections}

\section{@$\lambda$}
\begin{figure*}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} 
%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
\pipe e\\
		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
												\topp{\theta}{\theta}\\
\\
\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
%														\efix{x}{\tau}{e} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Tycon}{op}{
															\tauidx
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
												\ifix{\evar{x}}{\sigma}{\iota} \pipe
												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
												\iapp{\iota_{1}}{\iota_{2}} \pipe
												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
										& & \pipe & 
												\iunit \pipe
												\ipair{\iota_{1}}{\iota_{2}} \pipe 
												\ifst{\iota} \pipe
												\isnd{\iota} \pipe
												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \pipe \icase{e}{x}{e_1}{x}{e_2} \\

@@

\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
														\tlam{t}{\kappa}{\tau} \pipe 
														\tapp{\tau_1}{\tau_2} \pipe
														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
														\\

	 			& 		& \pipe	& 	 \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tlabel{label} \pipe \tunit \pipe 
														\tpair{\tau_{1}}{\tau_{2}} \pipe 
														\tfst{\tau} \pipe 
														\tsnd{\tau} 
														\\	
									&       & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \pipe \tsumcase{\tau}{x}{\tau_1}{x}{\tau_2} \\
\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
														\\													\text{types} 						& 		& \pipe	& 	\ttypestd \pipe \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\

\text{derivates} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype}  \pipe \tderbind{x}{t}{\tau}{\tau_1}\\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
 %& & \pipe & 
%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%														 \\

\text{reified IL}		&		&	\pipe	&	\titerm{\bar \iota} \pipe \titype{\bar \sigma} \\

@@

	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 

@@

	\pipe \cdots \pipe \iup{\tau} \pipe \tvalof{\tauiterm}{\tautype} \\						

@@

\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \kint \pipe

@@

@@

												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 

@@

												\kITerm \pipe \kIType

@@

\caption{\small Syntax of Core \atlam. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over the standard total binary operations over integers (e.g. addition). The form $\tvalof{\tauiterm}{\tautype}$ is  used internally by the semantics (it need not be supported by a parser).%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.

@@

\end{figure*}

@@

\label{z}&\quad\tops{z}{\kunit}{i}{a}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{
	\tapp{\tapp{\tvar{if\_empty}}{\tvar{args}}}{
		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}

@@

\label{s}&\quad\tops{s}{\kunit}{i}{a}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{\\
& \quad\quad
	\tapp{
		\tapp{\tvar{pop\_final}}{\tvar{args}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
	}}};\\
\label{rec}&\quad\tops{rec}{\kunit}{i}{args}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{\\
& \quad\quad
	\tapp{\tapp{\tvar{pop}}{\tvar{args}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{args}{\klist{\kDen}}{\\
	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{args}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{args}{\klist{\kDen}}{\\
	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{args}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
	)}}
	}}}}}}}}}}}}}}

@@

\label{nattype}&\etdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
& \elet{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat})})}}{

@@

}{\\
& \elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{}}}}{\\

@@

\vspace{-15px}

>>>>

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Typed programming languages are often described in fragments, each consisting of a small number of indexed type constructors (often one) and associated (term-level) operator constructors. The simply typed lambda calculus (STLC), for example, consists of a single fragment containing a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors: $\lambda$, indexed by a type, and \verb|ap|, which we may think of as being indexed trivially. A fragment is often identified by the type constructor it is organized around, so the STLC might also be called $\mathcal{L}\{\rightarrow\}$, following the notational convention used by Harper \cite{pfpl} and others.

@@

If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e. parametric polymorphism), we might take pause. In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined via Church-style encodings \cite{reynolds}\todo{what to cite for this?}. This fact is of substantial theoretical importance and is quite relevant if \emph{translating} a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to $\mathcal{L}\{\rightarrow\forall\}$ is one's main interest. But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, cautioned against using the existence of a weak encoding as an argument for programming with the corresponding embedding directly \cite{Reynolds94anintroduction}: 
\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
\end{quote}

%Adding new fragments directly to a language can make statically reasoning about programs more precise, support a more natural programming style and endow the language with a more favorable cost semantics. The latter two points might be relevant even if a strong embedding (i.e. one where there is a semantics-preserving isomorphism between terms and types of the fragment and the language) can be found. 
%
%Consistent with this view, typed programming languages like ML expose, for example, both $n$-ary product types and record types, despite the fact that any product type is definable using a record type (or in the other direction, using a module to provide the field selection operator). The compiler, on the other hand, \emph{is} mainly concerned with implementing the dynamic semantics correctly, and indeed, compilers often use the fact that weak definability results are quite readily derivable to their advantage, translating the constructs of an external language (EL) to those of a much simpler internal language (IL), e.g. with only binary products, during or directly after typechecking\todo{cite Harper-Stone}. %Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases needing consideration when implementing and verifying the compiler is a wise choice.

%Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, 
``General-purpose'' languages like ML, on the other hand, expose mechanisms that admit strong embeddings of many more fragments and occupy what is widely seen as a ``sweet spot'' in the design space. Nevertheless, situations continue to arise in both research and practice where introducing a new fragment would be desirable because an abstraction can still only be weakly defined in terms of existing fragments, or a strong embedding, while possible, is impractically verbose or inefficient:
\begin{enumerate}
\item General-purpose abstractions continue to evolve and admit variants. There are many  variations on record types, for example: records with functional record update, specified field ordering (that is, labeled tuples) or prototypic delegation are either awkward to strongly embed or cannot be. Similarly, sum types also admit variants (e.g. various forms of open sums). Even something as seemingly simple as a type-safe \verb|sprintf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.

\item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers. One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\item Foreign function interfaces (FFIs) that allow programmers to safely and naturally interact with libraries written in an existing language require enforcing the type system of the foreign language within the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
Safe FFIs generally require direct extensions to the language. For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{mlj}. For any other language, including  others on the JVM like Scala and the many languages that might be accessible via Java's native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
\end{enumerate}

These sorts of innovations are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, using tools like compiler generators, DSL frame\-works or  language workbenches, or directly within a particular compiler for the language, sometimes activated by a flag or pragma. This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined. Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so fragment providers have little choice but to leave this to clients.% We will consider this further in our discussion of related work in Sec. \ref{related-work}.

These are not the sorts of problems usually faced by library providers. Well-designed languages preclude the  possibility of conflict between two libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers. For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without needing to attend to burdensome ``link-time'' proof obligations.

\paragraph{Contributions}
We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define,  {using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language, from within. If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and logic.

The protocol constrains extensions to maintain type safety and decidability of type checking and type equality. Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively). The semantics guarantees that they will be conserved in the  ``open world'' by enforcing a form of  representation independence at all extension boundaries. Thus, if a strong embedding can be constructed using some combination of extensions, it can be constructed in the presence of arbitrary combinations of other extensions as well, avoiding the problems faced by language dialects and justifying the inclusion of the mechanism directly into the language.

We will begin in Sec. \ref{core} by introducing a minimal calculus with an actively typed semantics, @$\lambda$. The external language builds in only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products. We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes to make the guarantees that we claim. We then examine the semantics in greater detail and state and describe the key points in the proofs of the metatheory in Sec. \ref{theory}. These two sections contain the most fundamental contributions of this work.

While the core calculus addresses the issue of making a strong embedding possible, embeddings into the core calculus are quite clearly impractical syntactically. To begin to address the issue of syntax, we develop in Sec. \ref{expanded-syntax} a flexible type-directed dispatch protocol for a more conventional \emph{expanded syntax} that defers to the core calculus semantically. We then use the expanded syntax to discuss several more interesting examples in Sec. \ref{examples}: \todo{list these}. \todo{organization of rest of sections}.

While of surprising expressiveness, @$\lambda$ itself is certainly not capable of expressing the full variety of type systems listed above. Indeed, to guarantee safety and conservativity, the calculus is perhaps too conservative. For example, it only admits fragments for which the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment; new contexts cannot be introduced. We discuss this and some of the other reasons why a fragment may not be expressible in @$\lambda$ in Sec. \ref{limitations}. These limitations do not appear to be fundamental to the approach, so we also outline in Sec. \ref{limitations} promising directions for future work. We conclude with a discussion of related work in Sec. \ref{related-work}.

\section{Core @$\lambda$}\label{core}

%
%\begin{figure}[t]
%\begin{lstlisting}
%tycon Nat of 1 with  
%    schema $\lambda$idx:1.$\blacktriangledown$($\mathbb{Z}$) 
%    opcon Z of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].is_empty args $\llbracket$$\triangledown$(0) as Nat[()]$\rrbracket$)
%    opcon S of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].pop_final args $\lambda$x:ITm.$\lambda$ty:$\star$. 
%      checktype ty Nat[()] $\llbracket$$\triangledown$($\vartriangle$(x)+1) as Nat[()]$\rrbracket$)
%    opcon Rec of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].
%      pop args $\lambda$x1:ITm.$\lambda$ty1:$\star$.$\lambda$args':list[Elab].
%      pop args' $\lambda$x2:ITm.$\lambda$ty2:$\star$.$\lambda$args'':list[Elab].
%      pop_final args'' $\lambda$x3:ITm.$\lambda$ty3:$\star$.
%      check_type t1 Nat[()] (
%      check_type t3 Arrow[(Nat[()], Arrow[(t2, t2)])] 
%      	$\llbracket$$\triangledown$((fix f:$\mathbb{Z} \rightarrow$ rep(t2) is $\lambda$x:$\mathbb{Z}$.
%	        if x = 0 then $\vartriangle$(x2) else $\vartriangle$(x3) (x - 1) (f (x - 1))) $\vartriangle$(x1)) as t2$\rrbracket$))
%end
%\end{lstlisting}
%\caption{An implementation of primitive natural numbers as internal integers in @$\lambda$.}
%\label{nat-atlam}
%\end{figure}

The syntax of Core @$\lambda$ is given in Fig. \ref{grammar}. An example of a program defining type and operator constructors that can be used to strongly embed G\"odel's \textbf{T} in @$\lambda$ is given in Fig. \ref{nat}. We will discuss its semantics and how precisely the strong embedding seen being used in this example works as we go on. Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example). We will provide more sophisticated examples where this is less feasible later on. %It consists of:
%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.

\subsection{Programs}
\begin{figure}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}
\\& &  \pipe & 
%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
 e\\
		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
												\topp{\theta}{\theta}\\
\\
\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
%														\efix{x}{\tau}{e} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Tycon}{op}{
															\tauidx
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
												\ifix{\evar{x}}{\sigma}{\iota} \pipe
												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
												\iapp{\iota_{1}}{\iota_{2}} 
\\\text{integers}&&\pipe&												
												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
\text{products}										& & \pipe & 
												\iunit \pipe
												\ipair{\iota_{1}}{\iota_{2}} \pipe 
												\ifst{\iota} \pipe
												\isnd{\iota} 
\\\text{sums}&&\pipe&												
												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&&\pipe& \icase{e}{x}{e_1}{x}{e_2} \\

@@

\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
														\tlam{t}{\kappa}{\tau} \pipe 
														\tapp{\tau_1}{\tau_2} \pipe \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tlabel{label}
\\\text{lists}&&\pipe&											
														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
														\\

\text{products}	 			& 		& \pipe	& 	  \tunit \pipe 
														\tpair{\tau_{1}}{\tau_{2}} \pipe 
														\tfst{\tau} \pipe 
														\tsnd{\tau} 
														\\	
\text{sums}									&       & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\&&\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
						\text{types} 						& 		& \pipe	& 	\ttypestd \\&&\pipe& \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\

\text{derivates} 				& 		 & 	\pipe	&	\tden{{\bar\iota}}{\tau} \pipe \tden{\ibar}{\tau}^{\checkmark} \pipe \ttypeof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
 %& & \pipe & 
%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%														 \\

\text{rep types}		&		&	\pipe	&	\titype{\bar \sigma} \\

@@

\text{abstract IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 

@@

	\pipe \cdots \pipe \itransof{\tau} \\						

@@

\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \kint \pipe

@@

											    \klist{\kappa} \pipe

@@

												\kpair{\kappa_{1}}{\kappa_{2}} \\
												&&\pipe&

@@

												\kIType								

@@

\caption{\small Syntax of Core \atlam. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over standard total binary operations over integers (e.g. addition).%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.

@@

@@

\label{z}&\quad\tops{z}{\kunit}{i}{a}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{
	\tapp{\tapp{\tvar{arity0}}{\tvar{a}}}{\\&\quad\quad
		\tden{0}{\ttype{Nat}{\tunit}}

@@

\label{s}&\quad\tops{s}{\kunit}{i}{a}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{\tvar{arity1}~\tvar{a}~\tlam{d}{\kDen}{\\
& \quad\quad \tvar{ifeq}~\ttypeof{\tvar{d}}~\ttype{Nat}{\tunit}\\
& \quad\quad\quad \tden{\itransof{\tvar{d}} + 1}{\ttype{Nat}{\tunit}}
\label{rec}&\quad\tops{rec}{\kunit}{i}{args}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{\tvar{arity3}~\tvar{a}~\tlam{d1}{\kDen}{\tlam{d2}{\kDen}{\tlam{d3}{\kDen}{\\
&\quad\quad \tvar{ifeq}~\ttypeof{\tvar{d1}}~\ttype{Nat}{\tunit}\\
&\quad\quad {\sf let}~\tvar{t2} = \ttypeof{\tvar{d2}}~{\sf in}\\
&\quad\quad \tvar{ifeq}~\ttypeof{\tvar{d3}}~\ttype{Arrow}{(\ttype{Nat}{\tunit}, \ttype{Arrow}{(\tvar{t2}, \tvar{t2})})}\\
&\quad\quad\quad \tden{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
		\label{lastop}&\quad\quad\quad\quad\quad \iIfEq{x}{0}{\dint}{\itransof{\tvar{d2}}}{\\
		&\quad\quad\quad\quad\quad\quad\iapp{\iapp{\itransof{\tvar{d3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}}\\
		&\quad\quad\quad\quad)}{\itransof{\tvar{d1}}}}{\tvar{t2}}

@@

\label{nattype}&{\sf let}~\tvar{nat} = \ttype{Nat}{\tunit}~{\sf in}\\
& \elet{two}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{}}}}{\\
& \elet{plus}{

@@

}{
& \\

@@

%\vspace{-15px}
\caption{\small G\"odel's $\mathbf{T}$ in \atlam, used to calculate 2+2. The simple helper functions $\tvar{arity}n$ and $\tvar{ifeq}$ have return kind $\ksum{\kDen}{\kunit}$ (see text). Their definitions can be found in the appendix. We use \textsf{let} to bind both external and type-level variables for clarity of presentation; these can be eliminated manually. We will discuss the issue of a more natural syntax in Sec. \ref{expanded-syntax}.}
